The Count Distinct Problem

### 2015-11-05 wim/paper.tex
this paper is destroying my life

@@

%\\\\
%\indent The count-distinct problem has not always been a problem. 
%Before the internet was widely used by the population, storing data was expensive, and as a result most companies did not want to spend money on the hardware and space that would be required to store large amounts of data. 
%In addition, data was not transferred and gathered at a fast enough rate to warrant finding a way to count the data; even if companies wanted to count the data, the computers at the time were not powerful enough to perform the necessary calculations. 
%The problem of counting big data did not arise until the 1990's, when the internet began to facilitate the easy transfer and collection of data. 
%Today, more than 25 exabytes of data are created and stored a day. 
%The main factor that contributes most to the problem is that gathering and storing data is cheap, but analyzing it is not. The volume of the data leads to the count distinct problem that we will analyze \cite[pp. 1]{Yousra}.

@@

\indent The first solution to the count distinct problem is called the \textit{Hash Table}. 

@@

The following Java code can be used to model our hash table:\\\\
\begin{tabular}{c c} 
\begin{BVerbatim}
public int getHashValue(){
    char[] array = nameOfPokemon.toCharArray();
    int total = 0;
    for(char c : array){
	total += (int)c;
    }
    total += type;
    total %= 721;
    return total;
}
\end{BVerbatim}
\begin{BVerbatim}
...
for(Pokemon pm : S){
    int hashValue = pm.getHashValue();
    storeHashValue();
}
calculateAndPrintUniqueValues();
\end{BVerbatim}
Running this code will give us a final answer of 458 unique encounters with wild pok\'emon. 

@@

If we consider a larger data set, however, we may not be able to create a surjective hash function.

@@

\end{quote}

@@

Since the collision policy would not help us, we need to come up with a better hashing function. 
\indent More complex hash functions, such as SHA-2 (Secure Hash Algorithm 2) or MD5 (Message Digest 5) offer a solution to the collision problem. SHA-2 is used by organizations such as the NSA to encrypt data, and has yet to produce any collisions when hashing values \cite[p. 301]{SHA-Algo}. However, this is not a viable solution to the count distinct problem either due to the volume of the data being processed. In both the Twitter and Facebook problems, we are analyzing exabytes of data. The SHA-2 function outputs a 512 byte long integer for every input, which means that the resulting size of the set $\mathbb{V}$ would be in the worst case $512 \times ||\mathbb{S}||$. All of $\mathbb{V}$ would have to be held in memory, which means that analyzing 50 million hashtags (which is a low estimate)\cite{Twitter} could possibly result in $(50 \times 10^{10}) \times 512$ bytes of data being stored at one time, assuming every hashtag was 10 bytes long. A hash table of this size could not be loaded all at once into main memory; it would have to be stored into secondary memory which would exponentially increase the amount of time required to read the data \cite[p. 209]{Whang}. MD5 outputs smaller sized integers (128 in this case), but the collision probability is much higher \cite[p. 22 - 23]{Break-MD5}. The hash method is not a viable solution to the count distinct problem when we are dealing with a large volume of data, but hash functions are still used in the other solutions to the count-distinct problem.
\indent The hash-table method provided a 0\% error, but was not efficient in handling large sets of data. However, if we are looking for a \textit{good enough} estimate (within the neighborhood of 1\% to 5\% error) of the number of distinct elements in $\mathbb{S}$, we can use an algorithm called a Linear Probabilistic Counter. With the Linear Probabilistic Counter, the user can specify how large they want the error to be. This method uses a data structure called a bitmap. bitmaps are a matrix where either a 1 or 0 is stored in a cell, which allows for the map to be incredibly space efficient. The Linear Probabilistic Counter also uses a hash, but it only needs to temporarily produce a position and then it can ``forget'' the value it just produced, which in turn reduces the overhead memory needed to store the elements that was required by the hash table method \cite{Whang}. The number of distinct elements is determined by the number of 1's in the bitmap when the algorithm is finished.
\indent In order to either increase or decrease the error, the user of the Linear Probabilistic Counter algorithm can either increase or decrease the size of the bitmap that is held in main memory. The following equation gives us a relationship between the size of the bitmap and the error that will be produced:
bias\left(\frac{\hat{n}}{n}\right) = \frac{e^{t} - t - 1}{2n},
where $bias\left(\frac{\hat{n}}{n}\right)$ is the error. The value $t$ is the ratio of the number of distinct elements determined by the algorithm (denoted by $n$) to the size of the bitmap \cite{Whang}. If $t$ is close to 0, then the error will be very small; if $t$ is much greater than 1, the error will be very large. In other words, if the hash function being used produces a large amount of collisions, then the error will be high. Luckily, an algorithm such as SHA-2 will be okay to use, since the Linear Probabilistic Counter never stores the hashed value in main memory. We can also calculate the number of distinct elements by 

@@

where $m$ is the size of the bitmap and $\Lambda_{n}$ is the ratio of the number of 0's in the bitmap to the size of the bitmap \cite[p. 212]{Whang}. 
\indent The Facebook problem can be solved efficiently by using the Linear Probabilistic Counter algorithm. We would have set $\mathbb{S}$ of size 20 million \cite{Facebook}, and we can choose an arbitrary sized bitmap $\mathbb{V}$ of 10 million bits, or roughly 1.25 Mb, which lies on the upper limit of the efficiency of the Linear Probabilistic Counter. If the memory required for the bitmap increases beyond 1.25Mb, most operating systems will move the I/O operations to secondary memory, which defeats the purpose of the Linear Probabilistic Counter algorithm \cite[p. 211]{Whang}; as a result, the Twitter problem would not be efficiently solved by this method, since the amount of hashtags created in a day far exceeds 20 million. In the case of the Facebook problem, we can apply the Linear Probabilistic Counter algorithm analytically, rather than actually processing 20 million application installs.
%%%%%%%%FINISH THIS SECTION%%%%%%%%%
\indent We know that on a given day, Facebook users make 20 million app installs, and that there are currently about 7 million apps on Facebook \cite{Facebook}. 
However, not all 7 million applications are installed every day; in fact, the amount of installs decays exponentially starting from the number one most installed application on Facebook, which is Spotify as of August 2015 \cite{Facebook-2}. We would expect that apps such as Spotify would dominate the market, whereas apps in the lower spectrum of the downloads would be negligible. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

which allows the algorithm to increase in accuracy as the size of the original set grows \cite{Yousra}.

@@

We will use the code in \textit{Figure 1} to aggregate a random sampling of the most recent 2000 tweets that contain a ``\#'' every two minutes for 24 hours.

@@

@@

We know that the HyperLogLog can process nearly 1 billion elements with only 1.5 kilobytes of space to an accuracy of 2\% \cite{Flaj}.
\begin{floatingfigure}[r]{8cm}

\centering
\includegraphics[scale=0.3]{twitter_problem/ruby_code_01}
\caption{An infinite while loop can be useful}
\end{floatingfigure}
\noindent This is given by the equation we have for memory.
Using a similar argument, we can come up with the amount of memory we will need. 
If we want the error to be 1\%, then we can calculate the number of spaces required in the bitmap by using the equation we have for the error.
\begin{align*}
0.01 &= \frac{\sqrt{3\log{(2)} - 1}}{\sqrt{m}}\\
0.01\,\sqrt{m} &= \sqrt{3\log{(2)} - 1}\\
\sqrt{m}\, &= \frac{\sqrt{3\log{(2)} - 1}}{0.01}\\
m &= \frac{\sqrt{3\log{(2)} - 1}}{0.01^{2}}\\
m &\approx 10,794.4\ \text{spaces in the bitmap}.
\end{align*}
Now we can find the total memory required for our logarithm. If we let the program from \textit{Figure 1} run for 24 hours, at worst it will parse 1,440,000 hashtags.
We can now calculate the memory required by
\begin{align*}
\text{Memory}\, &= \log_{2}{(\log_{2}{(1,440,000)})}\\
&= 4.35457 \, \text{bytes per entry}.
\end{align*}
Multiplying this number by the total number of entires gives us a grand total of 47005 bytes, or 4.7005 kilobytes of required memory.
\\\\
\indent We now have all the necessary information to parse the data. We will implement the HyperLogLog in an object oriented manner; that is, we will create a Ruby class called HyperLogLog that we can initialize with the number of bytes per entry.
We also need a method to calculate the unique elements, and we need a way to actually call this routine.
\begin{center}
\begin{tabular}{c c c}
\includegraphics[scale=0.3]{twitter_problem/hll_init}
&
\includegraphics[scale=0.4]{twitter_problem/hll_count}
&
\includegraphics[scale=0.4]{twitter_problem/hll_imp}
\\
\textit{Initialization} & \textit{Counting} & \textit{Implementing}
\end{tabular}
\end{center}
\noindent Note that we settled with 4 as the size of each entry in the bit-map. This value is close enough to 4.35 to safely say that the error will be just about 1\%.
Finally, in our main program, we will loop infinitely (it will be up to us as to when to kill the process), and read in $\sim 2000$ tweets every two minutes. 
After 24 hours, the HyperLogLog algorithm outputs 
%%%%%%%%RESULTS GO HERE%%%%%%%%%%
\\\\
\indent

\newpage
\begin{thebibliography}{9}
\bibitem{Facebook} Facebook Statistics. (2015, September 20). Retrieved October 30, 2015, from \url{http://www.statisticbrain.com/facebook-statistics/}
\bibitem{Whang} Whang, Kyu-Young, Vander-Zanden, Brad T. \& Taylor, Howard M. (1990). A Linear-time Probabilistic Counting Algorithm for Database Applications. ACM Trans. Database Syst., 15, 208-229.
\bibitem{Maurer} Maurer, W. D. \& Lewis, T. G. (1975). Hash Table Methods. ACM Comput. Surv., 7, 5-19.
\bibitem{SHA-Algo} Chaves, Ricardo, Kuzmanov, Georgi, Sousa, Leonel \& Vassiliadis, Stamatis (2006). Improving SHA-2 Hardware Implementations. , 4249, 298-310.
\bibitem{Break-MD5} Wang, Xiaoyun \& Yu, Hongbo (2005). How to Break MD5 and Other Hash Functions. , 3494, 19-35.
\bibitem{Heule} Heule, S., Nunkesser, M., \& Hall, A. (2013). HyperLogLog in Practice: Algorithmic Engineering of a State of the Art Cardinality Estimation Algorithm. In \textit{Proceedings of the 16th International Conference on Extending Database Technology} (Vol. EDBT '13, pp. 683-692). New York, NY, Genoa: ACM.
\bibitem{Twitter} \#numbers. (2011, March 14). Retrieved October 29, 2015, from \url{https://blog.twitter.com/2011/numbers}
\bibitem{Pokemon} How many Pokemon are there altogether, before and after Omega Ruby and Alpha Sapphire? (2014, October 27). Retrieved October 29, 2015, from \url{http://pokemondb.net/pokebase/219332/pokemon-there-altogether-before-after-omega-alpha-sapphire}
\bibitem{Facebook-2} Most popular Facebook apps 2015 | Statistic. (2015, August 1). Retrieved November 3, 2015, from \url{http://www.statista.com/statistics/276371/most-popular-facebook-apps-by-monthly-active-users/}
\bibitem{Twitter-2} Twitter Statistics. (2015, September 15). Retrieved November 3, 2015, from \url{http://www.statisticbrain.com/twitter-statistics/}
\bibitem{Yousra}Yousra Chabchoub, Georges H\'ebrail. \textit{Sliding HyperLogLog: Estimating cardinality in a data stream.} 2010. 
\bibitem{Flaj} Flajolet, P., Fusy, \'E., Gandouet, O., \& Meunier, F. (n.d.). HyperLogLog: The Analysis of a Near-Optimal Cardinality Estimation Algorithm. 2007 \textit{Conference on Analysis of Algorithms, 07,} 127-143. Retrieved October 27, 2015, from \url{http://algo.inria.fr/flajolet/Publications/FlFuGaMe07.pdf}
\end{thebibliography}
\end{document}

>>>>

\usepackage{framed}

@@

@@

\indent The first solution to the count distinct problem is called a hash table. 

@@

\begin{floatingfigure}[H!]{8cm}
\centering
\begin{framed}
\includegraphics[scale=0.4]{pkmn_problem/hash_01}
\caption{Hashing Pok\'emon based on name and type}
\end{framed}
\end{floatingfigure}
\noindent Running the code in \textit{Figure 1} will give us a final answer of 461 unique encounters with wild pok\'emon. 

@@

If we consider a larger data set, however, we may not be able to create a surjective hash function.\\

@@

\end{quote}  \clearpage

@@

Since the collision policy would not help us, we need to come up with a better hashing function.\\
\indent More complex hash functions, such as SHA-2 (Secure Hash Algorithm 2) or MD5 (Message Digest 5) offer a solution to the collision problem. SHA-2 is used by organizations such as the NSA to encrypt data, and has yet to produce any collisions when hashing values \cite[p. 301]{SHA-Algo}. However, this is not a viable solution to the count distinct problem either due to the volume of the data being processed. In both the Twitter and Facebook problems, we are analyzing exabytes of data. The SHA-2 function outputs a 512 byte long integer for every input, which means that the resulting size of the set $\mathbb{V}$ would be in the worst case $512 \times ||\mathbb{S}||$. All of $\mathbb{V}$ would have to be held in memory, which means that analyzing 50 million hashtags (which is a low estimate)\cite{Twitter} could possibly result in $(50 \times 10^{10}) \times 512$ bytes of data being stored at one time, assuming every hashtag was 10 bytes long. A hash table of this size could not be loaded all at once into main memory; it would have to be stored into secondary memory which would exponentially increase the amount of time required to read the data \cite[p. 209]{Whang}. MD5 outputs smaller sized integers (128 in this case), but the collision probability is much higher \cite[p. 22 - 23]{Break-MD5}. The hash method is not a viable solution to the count distinct problem when we are dealing with a large volume of data, but hash functions are still used in the other solutions to the count-distinct problem.\\
\indent The hash-table method provided a 0\% error, but was not efficient in handling large sets of data. However, if we are looking for a \textit{good enough} estimate (within the neighborhood of 1\% to 5\% error) of the number of distinct elements in $\mathbb{S}$, we can use an algorithm called a Linear Probabilistic Counter. With the Linear Probabilistic Counter, the user can specify how large they want the error to be. This method uses a data structure called a bitmap. bitmaps are a matrix where either a 1 or 0 is stored in a cell, which allows for the map to be incredibly space efficient. The Linear Probabilistic Counter also uses a hash, but it only needs to temporarily produce a position and then it can ``forget'' the value it just produced, which in turn reduces the overhead memory needed to store the elements that was required by the hash table method \cite{Whang}. The number of distinct elements is determined by the number of 1's in the bitmap when the algorithm is finished.\\
\indent The LPC algorithm is divided into several steps. In the first step, the algorithm creates an arbitrarily sized bitmap in memory, and then hashes the values into the bitmap by determining the address in the bitmap at which the corresponding bit should be changed to a 1.
In the next step, the algorithm counts the number of 0's in the bitmap, and stores the result.
Finally, the algorithm estimates the cardinality of the unique set by dividing the count of 0's by the size of the bitmap, and then plugging in the result to

@@

where $m$ is the size of the bitmap and $\Lambda_{n}$ is the ratio \cite{Whang}. 
Figure 2 shows the Ruby implementation of the LPC algorithm.
\begin{floatingfigure}[R]{6cm}
\begin{framed}
\centering
\includegraphics[scale=0.4]{fb_problem/lpc}
\caption{Determining the cardinality}
\end{framed}
\end{floatingfigure}
\noindent The LPC algorithm also allows us to specify the amount of error we are willing to tolerate.
This is accomplished by changing the size of the bitmap; if the bitmap is large, or close to the size of the original set $\mathbb{S}$, then the error will be small.
Alternatively, if we decrease the size of the bitmap, we use less memory, but the error grows exponentially \cite{Hoff}.\\
\indent We can use the LPC algorithm to solve the Facebook problem; unfortunately, we have no way of accessing the actual data we need, so we will have to do some guessing. 
We know that Facebook users make about 7,000,000 app installs a dat \cite{Facebook-2}.
This means that the size of the set $\mathbb{S}$ will be 7,000,000.
Now we need to make some guesses; we expect that the applications that are most installed will appear a lot in our set, and that the applications that are installed less will appear less in $\mathbb{S}$. 
We do have access the to the top 20 Facebook application installs in August 2015, so we can plot those points and look for a trend in the data.
Doing so tells us that the data follows an exponential curve \cite{Facebook-2}, which means that Spotify (the top application at the time) will appear frequently in $\mathbb{S}$, whereas LINE, the $20^{\text{th}}$ most popular Facebook app, will appear sparingly.
This means that we will need a bitmap relatively close in size to the original set, since only a select few of the application have a high appearance rate in $\mathbb{S}$.
In other words, we expect a lot of distinct elements in $\mathbb{S}$.\\
\indent We can model the exponential function by choosing random numbers with a certain weight towards different numbers.
We will let the each number correspond to a different application.
For example, if we let 200 be the number for Spotify, we would expect a high appearance rate of 200 in $\mathbb{S}$.\\
%%%Need to figure out how to do this

@@

which allows the algorithm to increase in accuracy as the size of the original set grows \cite{Yousra}.\\

@@

We will use the code in \textit{Figure 2} to aggregate a random sampling of the most recent 2000 tweets that contain a ``\#'' every two minutes for 24 hours.

@@

\begin{floatingfigure}[hr]{8cm}
\centering
\begin{framed}
\includegraphics[scale=0.3]{twitter_problem/ruby_code_01}
\caption{An infinite while loop can be useful}
\end{framed}
\end{floatingfigure}

@@

\noindent We know that the HyperLogLog can process nearly 1 billion elements with only 1.5 kilobytes of space to an accuracy of 2\% \cite{Flaj}.\clearpage
This is given by the equation we have for memory.
Using a similar argument, we can come up with the amount of memory we will need. 
If we want the error to be 1\%, then we can calculate the number of spaces required in the bitmap by using the equation we have for the error.
\begin{align*}
0.01 &= \frac{\sqrt{3\log{(2)} - 1}}{\sqrt{m}}\\
0.01\,\sqrt{m} &= \sqrt{3\log{(2)} - 1}\\
\sqrt{m}\, &= \frac{\sqrt{3\log{(2)} - 1}}{0.01}\\
m &= \frac{\sqrt{3\log{(2)} - 1}}{0.01^{2}}\\
m &\approx 10,794.4\ \text{spaces in the bitmap}.
\end{align*}
Now we can find the total memory required for our logarithm. If we let the program from \textit{Figure 1} run for 24 hours, at worst it will parse 1,440,000 hashtags.
We can now calculate the memory required by
\begin{align*}
\text{Memory}\, &= \log_{2}{(\log_{2}{(1,440,000)})}\\
&= 4.35457 \, \text{bytes per entry}.
\end{align*}
Multiplying this number by the total number of entires gives us a grand total of 47005 bytes, or 4.7005 kilobytes of required memory.\\
\indent We now have all the necessary information to parse the data. We will implement the HyperLogLog in an object oriented manner; that is, we will create a Ruby class called HyperLogLog that we can initialize with the number of bytes per entry.
We also need a method to calculate the unique elements, and we need a way to actually call this routine.
\begin{tabular}{c c c}
\includegraphics[scale=0.3]{twitter_problem/hll_init}
\includegraphics[scale=0.4]{twitter_problem/hll_count}
\includegraphics[scale=0.4]{twitter_problem/hll_imp}
\\
\textit{Initialization} & \textit{Counting} & \textit{Implementing}
\noindent \\
Note that we settled with 4 as the size of each entry in the bit-map. This value is close enough to 4.35 to safely say that the error will be just about 1\%.
Finally, in our main program, we will loop infinitely (it will be up to us as to when to kill the process), and read in $\sim 2000$ tweets every two minutes. 
After running the algorithm for 24 hours, we will have a\\ 
%%%%%%%%RESULTS GO HERE%%%%%%%%%%
\indent We have looked at three different approaches to the count distinct problem. 
We used the HyperLogLog to solve only the Twitter problem, but we could have used the algorithm to solve all of the problems. 
The count distinct problem

\newpage
\begin{thebibliography}{9}
\bibitem{Hoff} Hoff, T. (2012, April 5). Big Data Counting: How to count a billion distinct objects using only 1.5KB of Memory - High Scalability -. Retrieved October 27, 2015, from \url{http://highscalability.com/blog/2012/4/5/big-data-counting-how-to-count-a-billion-distinct-objects-us.html}
\bibitem{Facebook} Facebook Statistics. (2015, September 20). Retrieved October 30, 2015, from \url{http://www.statisticbrain.com/facebook-statistics/}
\bibitem{Whang} Whang, Kyu-Young, Vander-Zanden, Brad T. \& Taylor, Howard M. (1990). A Linear-time Probabilistic Counting Algorithm for Database Applications. ACM Trans. Database Syst., 15, 208-229.
\bibitem{Maurer} Maurer, W. D. \& Lewis, T. G. (1975). Hash Table Methods. ACM Comput. Surv., 7, 5-19.
\bibitem{SHA-Algo} Chaves, Ricardo, Kuzmanov, Georgi, Sousa, Leonel \& Vassiliadis, Stamatis (2006). Improving SHA-2 Hardware Implementations. , 4249, 298-310.
\bibitem{Break-MD5} Wang, Xiaoyun \& Yu, Hongbo (2005). How to Break MD5 and Other Hash Functions. , 3494, 19-35.
\bibitem{Heule} Heule, S., Nunkesser, M., \& Hall, A. (2013). HyperLogLog in Practice: Algorithmic Engineering of a State of the Art Cardinality Estimation Algorithm. In \textit{Proceedings of the 16th International Conference on Extending Database Technology} (Vol. EDBT '13, pp. 683-692). New York, NY, Genoa: ACM.
\bibitem{Twitter} \#numbers. (2011, March 14). Retrieved October 29, 2015, from \url{https://blog.twitter.com/2011/numbers}
\bibitem{Pokemon} How many Pokemon are there altogether, before and after Omega Ruby and Alpha Sapphire? (2014, October 27). Retrieved October 29, 2015, from \url{http://pokemondb.net/pokebase/219332/pokemon-there-altogether-before-after-omega-alpha-sapphire}
\bibitem{Facebook-2} Most popular Facebook apps 2015 | Statistic. (2015, August 1). Retrieved November 3, 2015, from \url{http://www.statista.com/statistics/276371/most-popular-facebook-apps-by-monthly-active-users/}
\bibitem{Twitter-2} Twitter Statistics. (2015, September 15). Retrieved November 3, 2015, from \url{http://www.statisticbrain.com/twitter-statistics/}
\bibitem{Yousra}Yousra Chabchoub, Georges H\'ebrail. \textit{Sliding HyperLogLog: Estimating cardinality in a data stream.} 2010. 
\bibitem{Flaj} Flajolet, P., Fusy, \'E., Gandouet, O., \& Meunier, F. (n.d.). HyperLogLog: The Analysis of a Near-Optimal Cardinality Estimation Algorithm. 2007 \textit{Conference on Analysis of Algorithms, 07,} 127-143. Retrieved October 27, 2015, from \url{http://algo.inria.fr/flajolet/Publications/FlFuGaMe07.pdf}
\end{thebibliography}
\end{document}
